Improving modulation classification by distilling maximum likelihood \thanks{This work has been sponsored by armasuisse Science and Technology under the project ARNO (project code ARAMIS 047-22).}

### 2022-10-13 paper/main.tex
Add mar feedback

\begin{abstract}
Deep Neural Networks are being extensively used in communication systems and Automatic Modulation Classification (AMC) specifically, despite their susceptibility to adversarial perturbations, small carefully crafted noise that completely changes the network prediction. Recently, several works have shown the importance of robust models for security and interpretability, and methods like knowledge distillation and adversarial training have been proposed. In this work, we highlight the importance of the quality of the training data used by the model in its final robustness. We propose to use the Maximum Likelihood function to generate better training labels, that better teach the model to be uncertain under heavy channel corruptions. We show that this hybrid approach overcomes the limitations of Maximum Likelihood (e.g. requiring many priors) and it is quite effective at improving the robustness and the accuracy of neural networks. 
\end{abstract}

@@

% Maximum likelihood. 
%   (More details can be found in "Survey of automatic modulation classification techniques").
Maximum likelihood \cite{huan1995likelihood,dobre2007survey,hameed2009likelihood} is the analytical solution for the AMC task. It computes the likelihood function of the received signal with all possible modulations and estimates the most likely modulation. Despite being Bayes-optimal, because it can be computationally expensive and requires prior knowledge of the channel characteristics, sub-optimal approximations are used in practice \cite{dobre2007survey,Hameed_Dobre_Popescu_2009}.

@@

% Knowledge distillation
Knowledge distillation~\cite{hinton2015distilling,romero2014fitnets,zagoruyko2016paying,chebotar2016distilling} consists in training a student model to match the outputs of a teacher model, which has been shown to be effective in transferring performance. However, these approaches do not help the student network in benefitting from the higher adversarial robustness that a teacher network could have~\cite{goldblum2020adversarially}. However, recently, there are works that have proposed alternative loss functions that address this problem~\cite{goldblum2020adversarially,zi2021revisiting,zhu2021reliable,shao2021and,maroto2022benefits}. These solutions encourage and improve the student robustness by matching its output with the teacher outputs on adversarial examples as well as clean examples.

@@

While in many other tasks, these probabilities would be difficult or impossible to obtain, in AMC, we can use the maximum likelihood (ML) function to derive the real probabilities. Due to the characteristics of ML, this is only computationally feasible in coherent scenarios, where we have prior knowledge of the channel conditions. Particularly, if we assume the channel is gaussian with variance $\sigma^{2}$, and given the received symbols $r(x)$, where $r$ applies the proper communication filter and downsampling to the received signal. We show in Figure \ref{fig:comm_system} a diagram describing the communication system. We highlight that while the neural network has to process directly what the receptor receives, maximum likelihood processes the symbols of that signal instead. This is possible because we can assume we know the communication filter and upsampling used in the transmitter, which we can invert at the receptor. The ML function when the channel is gaussian is given by the following expression
\begin{equation}
	f_{i}(r(x)) = \dfrac{1}{2\pi \sigma^{2}|M_i|}\prod_{t=1}^{T} \sum_{j=1}^{|M_i|} \exp\left(-\dfrac{\lVert r(x)_t - s_j\rVert^{2}}{2 \sigma^{2}}\right)
\label{eq:ml}
\end{equation}
where $f_{i}$ is the ML function for a given modulation $M_i$ with $|M_i|$ number of states, $T$ is the number of symbols, $r(x)_t$ is the t-th symbol received, and $s_j$ is the symbol that correspond to the state $j$. Given a set of modulations, the probability of a modulation $M_i$ would be $p_{ML}^{(i)}(x) = \log(f_{i}(r(x))) / \sum_i \log(f_{i}(r(x)))$.

@@

\end{tikzpicture}
\caption{Simplified diagram of the communication system. The transmitter sends the binary data using the modulator, which converts the data into symbols, and $r'(x)$, which upsamples the signal and passes it through a communication filter (e.g. root raised cosine). The signal passes through the communication channel, which adds noise and distortion to the signal. The receptor does the inverse operation and demodulates the signal based on the neural network prediction. As shown, the signals used by the neural network and ML are slightly different.}
\label{fig:comm_system}
\end{figure}

@@

However, in practice, the neural network is trained with noisy signals where the modulation that was used cannot be inferred so easily. Intuitively, for these signals we would like the outputs of the network to rely on this sense of uncertainty. Thus, with similar principles as knowledge distillation, we propose to match the neural network outputs with the output of ML in the corresponding coherent scenario.
\begin{equation}
    \text{ST\_ML} : \dfrac{1}{N}\sum_{i}\Ls(f_{\theta}(x_i), p_{ML}(r(x_i)))
\label{eq:st_ml}
\end{equation}
where $p_{ML}$ is the Maximum Likelihood function, and $r$ is the function that uses the coupled receiver filter and downsamples the signal. Because $r$ and $p_{ML}$ are functions that depend on information that is not known in non-coherent scenarios, we propose to use this loss function only when training the network, where this information can be known.

@@

\begin{table}[htbp]
	\centering
	%\small
	\begin{tabular}{c|cc}
	    Method & Accuracy & Robustness \\
		\hline
		ST & $85.78 \pm 0.10$ & $20.56 \pm 1.14$ \\ 
		ST\_LNR & $86.21 \pm 0.08$ & $22.82 \pm 0.88$ \\ 
		ST\_ML & $\textbf{86.58} \pm 0.22$ & $\textbf{24.08} \pm 2.66$ \\
        \hline
		AT & $59.61 \pm 6.44$ & $55.45 \pm 3.64$ \\ 
		AT\_LNR & $64.53 \pm 1.76$ & $\textbf{59.05} \pm 1.42$ \\ 
		AT\_ML & $\textbf{65.06} \pm 1.77$ & $58.72 \pm 0.83$ \\ 
    \end{tabular}
    \caption{Accuracy and robustness for the ResNet model when varying the noise level, the samples per symbol, and the transmitter communication filter on a gaussian channel.}
    \label{tab:sawgn2p}
\end{table}

\begin{table}[htbp]
	\centering
	%\small
	\begin{tabular}{c|cc}
	    Method & Accuracy & Robustness \\
		\hline
		ST & $78.34 \pm 0.02$ & $18.86 \pm 1.83$ \\ 
		ST\_LNR & $78.08 \pm 0.04$ & $18.97 \pm 3.03$ \\ 
		ST\_ML & $\textbf{78.40} \pm 0.22$ & $\textbf{20.00} \pm 1.61$ \\
		\hline
		AT & $50.74 \pm 2.46$ & $\textbf{47.15} \pm 2.53$ \\ 
		AT\_LNR & $48.49 \pm 0.46$ & $45.62 \pm 0.57$ \\  
		AT\_NML & $42.11 \pm 4.66$ & $38.47 \pm 4.88$ \\  
		AT\_AML & $\textbf{54.10} \pm 0.57$ & $\textbf{46.94} \pm 0.17$ \\ 
    \end{tabular}
    \caption{Accuracy and robustness for the ResNet model when varying the channel properties and the level noise.}
    \label{tab:sp0c20}
\end{table}

>>>>

\begin{abstract}
Deep Neural Networks are being extensively used in communication systems and Automatic Modulation Classification (AMC) specifically, despite their susceptibility to adversarial perturbations, i.e. small carefully crafted noise that completely changes the network prediction. Recently, several works have shown the importance of robust models for security and interpretability, and methods like knowledge distillation and adversarial training have been proposed. In this work, we highlight the importance of the quality of the training data used by the model in its final robustness. We propose to use the Maximum Likelihood function to generate better training labels, that better teach the model to be uncertain under heavy channel corruptions. We show that this hybrid approach overcomes the limitations of Maximum Likelihood, like requiring the communication channel parameters as a prior, and it is quite effective at improving the robustness and the accuracy of neural networks. 
\end{abstract}

@@

% Maximum likelihood. 
%   (More details can be found in "Survey of automatic modulation classification techniques").
Maximum likelihood (ML) \cite{huan1995likelihood,dobre2007survey,hameed2009likelihood} is the analytical solution for the AMC task. It computes the likelihood function of the received signal with all possible modulations and estimates the most likely modulation. This method is Bayes-optimal, which means that it is the most accurate and adversarially robust model for AMC. However, it can be computationally expensive and requires prior knowledge of the channel characteristics, so sub-optimal approximations are used in practice \cite{dobre2007survey,Hameed_Dobre_Popescu_2009}.

@@

% Knowledge distillation
Knowledge distillation consists in training a student model to match the outputs of a teacher model, which has been shown to be effective in transferring performance~\cite{hinton2015distilling,romero2014fitnets,zagoruyko2016paying,chebotar2016distilling}. However, these approaches do not help the student network in benefitting from the higher adversarial robustness that a teacher network could have~\cite{goldblum2020adversarially}. However, recently, there are works that have proposed alternative loss functions that address this problem~\cite{goldblum2020adversarially,zi2021revisiting,zhu2021reliable,shao2021and,maroto2022benefits}. These solutions encourage and improve the student robustness by matching its output with the teacher outputs on adversarial examples as well as clean examples.

@@

While in many other tasks, these probabilities would be difficult or impossible to obtain, in AMC, we can use the ML function to derive the real probabilities. Due to the characteristics of ML, this is only computationally feasible in coherent scenarios, where we have prior knowledge of the channel conditions. Particularly, if we assume the channel is gaussian with variance $\sigma^{2}$, and given the received symbols $r(x)$, where $r$ applies the proper communication filter and downsampling to the received signal. We show in Figure \ref{fig:comm_system} a diagram describing the communication system. We highlight that while the neural network has to process directly what the receptor receives, maximum likelihood processes the symbols of that signal instead. This is possible because we can assume we know the communication filter and upsampling used in the transmitter, which we can invert at the receptor. The ML function when the channel is gaussian is given by the following expression
\begin{equation}
	f_{i}(r(x)) = \dfrac{1}{2\pi \sigma^{2}|M_i|}\prod_{t=1}^{T} \sum_{j=1}^{|M_i|} \exp\left(-\dfrac{\lVert r(x)_t - s_j\rVert^{2}}{2 \sigma^{2}}\right)
\label{eq:ml}
\end{equation}
where $f_{i}$ is the ML function for a given modulation $M_i$ with $|M_i|$ number of states, $T$ is the number of symbols, $r(x)_t$ is the t-th symbol received, and $s_j$ is the symbol that correspond to the state $j$. Given a set of modulations, the probability of a modulation $M_i$ would be $p_{ML}^{(i)}(x) = \log(f_{i}(r(x))) / \sum_i \log(f_{i}(r(x)))$.

@@

\end{tikzpicture}
\caption{Simplified diagram of the communication system. The transmitter sends the binary data using the modulator, which converts the data into symbols, and $r'(x)$, which upsamples the signal and passes it through a communication filter (e.g. root raised cosine). The signal passes through the communication channel, which adds noise and distortion to the signal. The receptor uses a coupled communication filter with downsampling, $r(x)$, and demodulates the signal based on the neural network prediction. As shown, the signals used by the neural network and ML are slightly different.}
\label{fig:comm_system}
\end{figure}

@@

However, in practice, the neural network is trained with noisy signals where the modulation that was used cannot be inferred so easily. Intuitively, for these signals we would like the outputs of the network to rely on this sense of uncertainty. Thus, with similar principles as knowledge distillation, we propose to match the neural network outputs with the output of ML in the corresponding coherent scenario.
\begin{equation}
    \text{ST\_ML} : \dfrac{1}{N}\sum_{i}\Ls(f_{\theta}(x_i), p_{ML}(r(x_i)))
\label{eq:st_ml}
\end{equation}
where $p_{ML}$ is the Maximum Likelihood function, and $r$ is the function that uses the coupled receiver filter and downsamples the signal. We propose to use this loss function only during training, since $r$ and $p_{ML}$ are functions that require information that is unavailable in non-coherent scenarios (i.e. during test).

@@

\begin{table}[htbp]
	\centering
	%\small
	\begin{tabular}{c|cc}
	    Method & Accuracy & Robustness \\
		\hline
		ST & $85.78 \pm 0.10$ & $20.56 \pm 1.14$ \\ 
		ST\_LNR & $86.21 \pm 0.08$ & $22.82 \pm 0.88$ \\ 
		ST\_ML & $\textbf{86.58} \pm 0.22$ & $\textbf{24.08} \pm 2.66$ \\
        \hline
		AT & $59.61 \pm 6.44$ & $55.45 \pm 3.64$ \\ 
		AT\_LNR & $65.03 \pm 1.23$ & $59.20 \pm 1.34$ \\ 
		AT\_ML & $\textbf{66.27} \pm 1.72$ & $\textbf{59.74} \pm 1.29$ \\ 
    \end{tabular}
    \caption{Accuracy and robustness for the ResNet model when varying the noise level, the samples per symbol, and the transmitter communication filter on a gaussian channel.}
    \label{tab:sawgn2p}
\end{table}

\begin{table}[htbp]
	\centering
	%\small
	\begin{tabular}{c|cc}
	    Method & Accuracy & Robustness \\
		\hline
		ST & $78.34 \pm 0.02$ & $18.86 \pm 1.83$ \\ 
		ST\_LNR & $78.08 \pm 0.04$ & $18.97 \pm 3.03$ \\ 
		ST\_ML & $\textbf{78.40} \pm 0.22$ & $\textbf{20.00} \pm 1.61$ \\
		\hline
		AT & $50.74 \pm 2.46$ & $\textbf{47.15} \pm 2.53$ \\ 
		AT\_LNR & $48.49 \pm 0.46$ & $45.62 \pm 0.57$ \\ 
		AT\_ML & $49.01 \pm 0.13$ & $45.07 \pm 0.56$ \\  
		AT\_AML & $\textbf{54.10} \pm 0.57$ & $\textbf{46.94} \pm 0.17$ \\ 
    \end{tabular}
    \caption{Accuracy and robustness for the ResNet model when varying the channel properties and the level noise.}
    \label{tab:sp0c20}
\end{table}
